---
title: "Host-Pathogen Omics Explorer — Analysis Report"
author: "Bioinformatics Toolkit"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(dplyr)
library(knitr)
```

## 1. Database Overview

The analysis is built on a curated SQLite database of host-pathogen interactions,
covering 5 intracellular pathogens and 32 host proteins involved in vesicular
trafficking and phagosome maturation.

```{r db-overview}
# Count records from the database
library(RSQLite)

db_file <- file.path("..", "src", "hostpathogen", "data", "hostpathogen.db")
con <- dbConnect(SQLite(), db_file)

tables <- c("pathogens", "host_proteins", "effectors", "effector_targets", "maturation_stages")
for (t in tables) {
  count <- dbGetQuery(con, paste("SELECT COUNT(*) AS n FROM", t))$n
  cat("- **", t, "**: ", count, " rows\n", sep = "")
}

# Pathogen strategies
strategies <- dbGetQuery(con, "SELECT name, strategy FROM pathogens")
dbDisconnect(con)
```

```{r strategy-table}
kable(strategies, col.names = c("Pathogen", "Strategy"), caption = "Pathogen Evasion Strategies")
```

## 2. Differential Expression

Simulated RNA-seq data was analysed with DESeq2 to identify host genes
differentially expressed during infection (infected vs control, 3 samples each).

```{r de-summary}
res_file <- file.path("..", "r", "data", "deseq2_results.csv")
if (file.exists(res_file)) {
  res <- read.csv(res_file, row.names = 1)
  n_sig <- sum(res$padj < 0.05, na.rm = TRUE)
  cat("Total genes tested:", nrow(res), "\n")
  cat("Significant (padj < 0.05):", n_sig, "\n")
  cat("Upregulated (log2FC > 1):", sum(res$padj < 0.05 & res$log2FoldChange > 1, na.rm = TRUE), "\n")
  cat("Downregulated (log2FC < -1):", sum(res$padj < 0.05 & res$log2FoldChange < -1, na.rm = TRUE), "\n")
}
```

### Volcano Plot

```{r volcano, fig.width=8, fig.height=6}
volcano_file <- file.path("..", "r", "plots", "volcano_plot.png")
if (file.exists(volcano_file)) knitr::include_graphics(volcano_file)
```

### Heatmap of Top DE Genes

```{r heatmap, fig.width=8, fig.height=8}
heatmap_file <- file.path("..", "r", "plots", "heatmap.png")
if (file.exists(heatmap_file)) knitr::include_graphics(heatmap_file)
```

## 3. Network Analysis

The host-pathogen interaction network was built from curated effector-target
relationships. Centrality analysis identifies which host proteins are most
frequently targeted by pathogen effectors.

```{r hubs, fig.width=8, fig.height=5}
hub_file <- file.path("..", "r", "plots", "hub_targets.png")
if (file.exists(hub_file)) knitr::include_graphics(hub_file)
```

```{r hub-table}
net_file <- file.path("..", "r", "data", "network_metrics.csv")
if (file.exists(net_file)) {
  hubs <- read.csv(net_file)
  kable(head(hubs, 5), caption = "Top 5 Hub Host Proteins")
}
```

## 4. Pathway Enrichment

Pathway over-representation analysis identifies host pathways that are
enriched among DE genes or pathogen-targeted proteins.

```{r enrichment, fig.width=7, fig.height=5}
enrich_file <- file.path("..", "r", "data", "enrichment_results.csv")
if (file.exists(enrich_file)) {
  enrich <- read.csv(enrich_file)
  kable(head(enrich, 5), caption = "Top Enriched Pathways")
}

enrich_plot <- file.path("..", "r", "data", "enrichment_plot.png")
if (file.exists(enrich_plot)) knitr::include_graphics(enrich_plot)
```

## 5. Effector Repertoire

```{r effectors, fig.width=8, fig.height=4}
effector_plot <- file.path("..", "r", "plots", "effector_counts.png")
if (file.exists(effector_plot)) knitr::include_graphics(effector_plot)
```

## 6. Methods Summary

| Component | Technology | Purpose |
|-----------|-----------|---------|
| Data storage | SQLite | Curated host-pathogen interaction database |
| Data access | Python (sqlite3 + pandas) | Load, query, transform |
| Network analysis | Python (NetworkX) | Interaction graph, hub detection |
| Statistics | Python (scipy), R (DESeq2) | Differential expression, enrichment |
| Machine learning | Python (scikit-learn) | Evasion strategy prediction |
| Visualisation | R (ggplot2, pheatmap) | Publication-quality figures |
| Reporting | R Markdown | Reproducible analysis report |

---

*Generated by the Host-Pathogen Omics Explorer toolkit.*
